Sediment Interception by Emergent Stems Across Varying Patch Densities and Flows

### 2021-01-04 aa workingdraft/draft.tex
moving to R to fix turbulent vs turbulence in figure 5

Aquatic plants affect sedimentation via numerous mechanisms. Sediment is captured directly on stems, leaves, and other plant surfaces \cite{mudd2010does, peruzzo2012capillary}, and effects of plants on turbulence and flow patterns impact the rate at which suspended particles settle out of the water column due to gravity \cite{christiansen2000flow, leonard1995flow, Nielsen_1993, Jacobs_2016, Wang_2018}. Furthermore, plants affect bedload transport \cite{yager2013influence, yang2019impact, jordanova2003experimental} and reentrainment of cohesive \cite{d2007landscape} and sandy \cite{tinoco2018turbulence} sediment by turbulent flow, which both also impact long-term bed elevation change \cite{wu2005depth,d2007landscape}. While gravitational settling plays a dominant role for large or dense particles in typical flow conditions for vegetated reaches \cite{mudd2010does, leonard1995flow}, finer and lighter particles may be destined to either impact vegetation and be captured directly or pass through vegetated areas without being retained.

@@

Physical properties of collectors, particles, and flow environments are all known to affect capture efficiency, though the relative importance of different mechanisms of effect is still being determined. Greater flow velocities ($u$) and larger particle diameters ($d_p$) result in greater direct capture and inertial impaction of particles on stems \cite{Palmer_2004,fuchs1965mechanics}. However, increased $u$ results in increased turbulence---often parameterized by collector Reynolds number ($\Rey_c=\frac{ud_c}{\nu}$), where $\nu$ is kinematic viscosity---which can trap particles in vortices distant from collector surfaces and shear particles off of collectors, resulting in lower ECE. Increased $d_c$ and spatial density of collectors, which we parameterize as solid volume fraction ($\phi_c$), can also result in increased turbulence intensity due to collector wakes, although at sufficiently high densities, decreasing interstitial distance can limit turbulence scale and have a damping effect on turbulent kinetic energy (TKE) \cite{nepf_drag_1999}. Lastly, biofilm comprising algae, cyanobacteria, and other microorganisms that live on the surfaces of submerged vegetation, has been confirmed to affect capture efficiency \cite{Fauria_2015}. Increased stem roughness \cite{Palmer_2004} seems a likely mechanism, along with possibly improved particle adherence \cite{wu2014colloid}.

@@

\noindent where $C$, $a$, and $b$ are empirically determined regression coefficients. They used an isolated single cylinder coated with grease, for which retention was assumed to be complete ($p_r = 1$), and yielded a model,

@@

We conducted our experiments in the Ecogeomorphology flume, an indoor (22.2 +/- 1 \SI{}{\celsius}) recirculating flume located in McCone Hall at the University of California, Berkeley. The flume has a rectangular open-channel section (5.25 m L $\times$ 0.6 m W $\times$ 0.6 m H) with a bed and sidewalls that are smooth and transparent. At both its upstream and downstream ends, this section connects to rectangular ducts with gradually changing hydraulic diameter and rounded corners with curved vertical manifolds. These features are intended to standardize flow outside the open channel where experimental treatments were installed. Additionally, at the upstream end of the open channel, a honeycomb flow collimator further straightens flow (Figure \ref{fig:floorplan}).

@@

Within the open channel, we designated a "test section" 1.95 m in length where collectors would be installed. This test section and the surrounding area were instrumented with the following devices:
\begin{enumerate}
     \item A flat-bedded array positioned flush with the neighboring channel bed, containing vertical, emergent collector stems, which were spaced uniformly in an unpatterned manner
    \item Two battery-operated peristaltic pumps (Cole-Parmer, Vernon Hills, IL) used to sample suspended particle concentration via three hose inlets each (inside diameter = 3.1 mm), which were suspended at a range of heights from the channel bed (5; 14; 27 cm); these were positioned 50 cm upstream and downstream of the test section 
   \item sediment traps (n = 9) with 2.5 cm circular openings flush with the bed and collection filters (Whatman GF/F) on perforated filter holders recessed in a 5 cm deep cylindrical cavity (trap Reynolds number and aspect ratio chosen to minimize bias \cite{butman1986sediment}); these were interspersed among the collectors in a grid-like pattern
\end{enumerate}

We used 1/8" (0.3175 cm) cylindrical wooden dowels as collectors, which were covered in silicone grease (Chemplex 710, Fuchs Petrolub, Mannheim, Germany) in order to retain impacted particles and spread randomly throughout the test section at the appropriate density. Crushed walnut shell was used as a substitute for suspended sediment because it has surficial and physical properties comparable to the organic-rich types common in wetlands \cite{muller2017experiments, jenzer2015sediment, redding2006particle}. We used WF5-200 grade (Composition Materials Co., Milford, CT), which passes entirely through a \#60 sieve (\SI{250}{\micro\metre}), and which we determined has a average particle diameter ($d_{50}$) of \SI{25.2}{\micro\metre} based on measurements using a laser-scattering-based instrument (LISST-Portable|XR, Sequoia Scientific, Bellevue, WA; See Appendix A1). We also empirically estimated the particle density for this specific walnut shell flour to be 1.53 \SI{}{\gram/\centi\metre\cubed} using volume displacement.

@@

Sediment traps were installed immediately before each run, then a slurry consisting of 200 g of crushed walnut shell suspended in 15.1 L of tap water was added using a spigot calibrated to finish draining after a period roughly equal to circulation time (3 minutes). Test-section flow velocity was kept at 6 cm/s for this period across $\Rey_c$ treatments, then adjusted to the appropriate value for the treatment. This procedure was intended to make particle concentration longitudinally even throughout the system. Estimated depth-averaged starting concentration was 82.3 mg/L (\SI{53.8}{\micro\liter/\liter}).

In preliminary experiments, we determined 100 minutes was more than adequate to capture the effects of settling and capture on suspended concentrations across our parameter space. We sampled suspended concentration using the peristaltic pumps every 5 minutes, the first sample occurring 5 minutes after we began adding sediment to the flume. Samples measured approximately 140 ml each, and our sampling frequency resulted in 19 sampled timesteps per experiment, with six samples per timestep. sediment traps were covered with a plunger and removed at 100 minutes, so the last peristaltic pump sample occurred at 95 minutes due to personnel limitations.

@@

We estimated mid-water-column turbulent kinetic energy (TKE) among the collectors for each combination of $\Rey_c$ and collector density using the Vectrino Profiler ADV. The Vectrino was mounted on a frame suspended above the test section and adjusted until its bottom-depth measurement read 15 cm. A bubble level was used to ensure it was oriented along the vertical axis. The ADV measured velocity in two horizontal dimensions, and two separate measurements along the vertical dimension, at a frequency of 10 Hz. It also collected other data such as autocorrelation and signal-to-noise ratio. We collected 5 minutes of data per treatment in the interest of including the vast majority of expected periodicity in our variance measurement. Measurements taken by the ADV were filtered, leaving out timepoints where the two vertical velocity measurements differed by 50\% or more or the SNR measurement was greater than 5 dB, and were subsequently despiked using a phase-space thresholding algorithm \cite{goring_nikora}. TKE was derived from the cleaned Vectrino data as half the sum of the variances of the velocity components:
\begin{equation}
    \text{TKE} = \frac{1}{2}[\overline{(u^\prime)^2} + \overline{(v^\prime)^2} + \overline{(w^\prime)^2}]\,.
    \label{eqn:TKE}
\end{equation}

@@

% Turbulence table
\begin{table}[H]
\caption{Turbulent kinetic energy (TKE) for all collector solid volume fraction ($\phi_c$) and $\Rey_c$ treatments, including the control (zero-collector) treatments for reference.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{>{\bfseries}r>{\bfseries}rrr}
\toprule
\textbf{$\phi_c$}&\textbf{$\Rey_c$}&\textbf{mid-water-column TKE (\SI{}{\milli\metre^2/\second^2})}\\
\midrule
Control &   67  & \num{2.48}\\
        &   134 & \num{2.98}\\
        &   200 & \num{3.37}\\
\midrule
0.22\% &   67  & \num{1.23}\\
        &   134 & \num{5.26}\\
        &   200 & \num{12.2}\\
\midrule
0.64\% &   67  & \num{9.13}\\
        &   134 & \num{35.9}\\
        &   200 & \num{59.6}\\
\midrule
1.17\%  &   67  & \num{11.1}\\
        &   134 & \num{29.8}\\
        &   200 & \num{54.4}\\
\bottomrule
\label{tbl:turbulence}
\end{tabular}
\end{table}

\begin{figure}[H]
\centering
\includegraphics[width=4in]{../pics/tke.png}
\caption{Effective capture efficiency as a function of turbulent kinetic energy for all experimental treatments studied.}
\label{fig:tke}
\end{figure}

@@

The threshold in TKE's relationship with collector density likely indicates that the turbulent eddy scale and the distance between collectors converged somewhere in our parameter space, so that additional collectors acted to damp eddies as much or more than they acted to create them. Accounting for secondary turbulence in future models and experiments seems feasible and important. In general, moving on from the simplistic power-law models developed initially as a way to bridge the gap between analytical expressions for creeping and potential flows to more theoretically-grounded empirical models \cite{stein2020} seems a promising avenue for future investigation.

>>>>

Aquatic plants affect sedimentation via numerous mechanisms. Sediment is captured directly on stems, leaves, and other plant surfaces \cite{mudd2010does, peruzzo2012capillary}, and effects of plants on turbulence and flow patterns impact the rate at which suspended particles settle out of the water column due to gravity \cite{christiansen2000flow, leonard1995flow, Nielsen_1993, Jacobs_2016, Wang_2018}. Furthermore, plants affect bedload transport \cite{yager2013influence, yang2019impact, jordanova2003experimental} and reentrainment of cohesive \cite{d2007landscape} and sandy \cite{tinoco2018turbulence} sediment by turbulent flow, which both also impact long-term bed elevation change \cite{wu2005depth,d2007landscape}. While gravitational settling plays a dominant role for large particles in typical flow conditions for vegetated reaches \cite{mudd2010does, leonard1995flow}, finer particles may be destined to either impact vegetation and be captured directly or pass through vegetated areas without being retained.

@@

Physical properties of collectors, particles, and flow environments are all known to affect capture efficiency, though their relative importances are still unclear. Greater flow velocities ($u$) and larger particle diameters ($d_p$) result in greater direct capture and inertial impaction of particles on stems \cite{Palmer_2004,fuchs1965mechanics}. However, increased $u$ results in increased turbulence---often parameterized by collector Reynolds number ($\Rey_c=\frac{ud_c}{\nu}$), where $\nu$ is kinematic viscosity---which can trap particles in vortices distant from collector surfaces and shear particles off of collectors, resulting in lower ECE. Increased $d_c$ and spatial density of collectors, which we parameterize as the fraction of underwater volume in a patch of vegetation occupied by plant matter ($\phi_c$), can also result in increased turbulence intensity due to collector wakes. However, at sufficiently high densities, decreasing interstitial distance can limit turbulent eddy scale and have a damping effect on turbulence kinetic energy (TKE) \cite{nepf_drag_1999, leonard1995flow}. Lastly, biofilm comprising algae, cyanobacteria, and other microorganisms that live on the surfaces of submerged vegetation, has been confirmed to affect capture efficiency \cite{Fauria_2015}. Increased stem roughness \cite{Palmer_2004} seems a likely mechanism, along with possibly improved particle adherence \cite{wu2014colloid}.

@@

\noindent where $C$, $a$, and $b$ are empirically determined regression coefficients, and $R$ is the ratio between $d_p$ and $d_c$. They used an isolated single cylinder coated with grease, for which retention was assumed to be complete ($p_r = 1$), and yielded a model,

@@

We conducted our experiments in the Ecogeomorphology flume, an indoor (22.2 +/- 1 \SI{}{\celsius}) recirculating flume located in McCone Hall at the University of California, Berkeley. The flume has a rectangular open-channel section (5.25 m L $\times$ 0.6 m W $\times$ 0.6 m H) with a bed and sidewalls that are smooth and transparent. At both its upstream and downstream ends, this section connects to rectangular ducts with gradually changing hydraulic diameter and rounded corners with curved vertical manifolds. These features are intended to standardize flow outside the open channel where experimental treatments were installed. Additionally, at the upstream end of the open channel, a honeycomb flow collimator further straightens flow streamlines (Figure \ref{fig:floorplan}).

@@

Within the open channel, we designated a "test section" 1.95 m in length where collectors were installed. This test section and the surrounding area were instrumented with the following devices:
\begin{enumerate}
     \item A flat-bedded array positioned flush with the neighboring channel bed, containing vertical, emergent collector stems, which were spaced uniformly in an unpatterned manner
    \item Two battery-operated peristaltic pumps (Cole-Parmer, Vernon Hills, IL) used to sample suspended particle concentration via three hose inlets each (inside diameter = 3.1 mm), which were suspended at a range of heights from the channel bed (5; 14; 27 cm); these were positioned 50 cm upstream and downstream of the test section 
   \item Sediment traps (n = 9) with 2.5 cm circular openings flush with the bed and collection filters (Whatman GF/F) on perforated filter holders recessed in a 5 cm deep cylindrical cavity (trap Reynolds number and aspect ratio chosen to minimize bias \cite{butman1986sediment}); these were interspersed among the collectors in a grid-like pattern
\end{enumerate}

We used 1/8" (0.3175 cm) cylindrical wooden dowels as collectors, which were covered in silicone grease (Chemplex 710, Fuchs Petrolub, Mannheim, Germany) in order to retain impacted particles and spread randomly throughout the test section at the appropriate density. Crushed walnut shell was used as a substitute for suspended sediment because it has physical properties comparable to the organic-rich types common in wetlands \cite{muller2017experiments, jenzer2015sediment, redding2006particle}. We used WF5-200 grade (Composition Materials Co., Milford, CT), which passes entirely through a \#60 sieve (\SI{250}{\micro\metre}) and has an average particle diameter ($d_{50}$) of \SI{25.2}{\micro\metre} based on measurements using a laser-scattering-based instrument (LISST-Portable|XR, Sequoia Scientific, Bellevue, WA; See Appendix A1). We also empirically estimated the particle density of this specific walnut shell flour to be 1.53 \SI{}{\gram/\centi\metre\cubed} using volume displacement.

@@

Sediment traps were installed immediately before each run, then a slurry consisting of 200 g of crushed walnut shell suspended in 15.1 L of tap water was added using a spigot calibrated to finish draining after a period roughly equal to circulation time (3 minutes). Test-section flow velocity was kept at 6 cm/s during this period across $\Rey_c$ treatments, then adjusted to the appropriate value for the treatment. This procedure was intended to make particle concentration longitudinally even throughout the system. The estimated depth-averaged starting concentration was 82.3 mg/L (\SI{53.8}{\micro\liter/\liter}).

In preliminary experiments, we determined 100 minutes was more than adequate to capture the effects of settling and capture on suspended concentrations across our parameter space. We sampled suspended concentration using the peristaltic pumps every 5 minutes, the first sample occurring 5 minutes after we began adding sediment to the flume. Samples measured approximately 140 ml each, and our sampling frequency resulted in 19 sampled timesteps per experiment, with six samples per timestep. Sediment traps were covered with a plunger and removed at 100 minutes.

@@

We estimated mid-water-column turbulence kinetic energy (TKE) among the collectors for each combination of $\Rey_c$ and collector density using the Vectrino Profiler ADV. The Vectrino was mounted on a frame suspended above the test section and adjusted until its bottom-depth measurement read 15 cm. A bubble level was used to ensure it was oriented along the vertical axis. The ADV measured longitudinal, lateral, and vertical flow velocity vector components at a frequency of 10 Hz, plus an additional, independent duplicate of the vertical component for quality control \cite{nortek2018}. We collected 5 minutes of data per treatment in order to capture periodic fluctuations in flow and turbulence patterns. Measurements taken by the ADV were filtered to leave out timepoints where the two vertical velocity measurements differed by more than 50\% or the signal-to-noise-ratio reading was greater than 5 dB, and were subsequently despiked using a phase-space thresholding algorithm \cite{goring_nikora}. TKE was derived from the cleaned Vectrino data as half the sum of the variances of the velocity components:
\begin{equation}
    \text{TKE} = \frac{1}{2}[\overline{(u^\prime)^2} + \overline{(v^\prime)^2} + \overline{(w^\prime)^2}]\,.
    \label{eqn:TKE}
\end{equation}

@@

% Turbulence table
\begin{table}[H]
\caption{Turbulence kinetic energy (TKE) for all collector solid volume fraction ($\phi_c$) and $\Rey_c$ treatments, including the control (zero-collector) treatments for reference.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{>{\bfseries}r>{\bfseries}rrr}
\toprule
\textbf{$\phi_c$}&\textbf{$\Rey_c$}&\textbf{mid-water-column TKE (\SI{}{\milli\metre^2/\second^2})}\\
\midrule
Control &   67  & \num{2.48}\\
        &   134 & \num{2.98}\\
        &   200 & \num{3.37}\\
\midrule
0.22\% &   67  & \num{1.23}\\
        &   134 & \num{5.26}\\
        &   200 & \num{12.2}\\
\midrule
0.64\% &   67  & \num{9.13}\\
        &   134 & \num{35.9}\\
        &   200 & \num{59.6}\\
\midrule
1.17\%  &   67  & \num{11.1}\\
        &   134 & \num{29.8}\\
        &   200 & \num{54.4}\\
\bottomrule
\label{tbl:turbulence}
\end{tabular}
\end{table}

\begin{figure}[H]
\centering
\includegraphics[width=4in]{../pics/tke.png}
\caption{Effective capture efficiency as a function of turbulence kinetic energy for all experimental treatments studied.}
\label{fig:tke}
\end{figure}

@@

The threshold in TKE's relationship with collector density likely indicates that the turbulent eddy scale and the distance between collectors converged somewhere in our parameter space, so that additional collectors acted to damp eddies as much or more than they acted to create them. This would be consistent with previous \textit{in situ} evidence of a negative relationship between collector density and turbulence intensity \cite{leonard1995flow}. Accounting for secondary turbulence in future models and experiments seems feasible and important. In general, moving on from the simplistic power-law models developed initially as a way to bridge the gap between analytical expressions for creeping and potential flows to more theoretically-grounded empirical models \cite{stein2020} seems a promising avenue for future investigation.

### 2021-01-04 aa workingdraft/preamble.tex
updates
just write the damn discussion already
done for tonight
moving to R to fix turbulent vs turbulence in figure 5

\newcommand\Rey{\mathrm{Re}}

@@

\usepackage{threeparttable}

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{Suspended Particle Interception by Emergent Stems Across Varying Patch Densities and Flows}

@@

%TODO
% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{TO DO}

>>>>

\newcommand\Rey{\mathrm{Re}}
\newcommand\Pec{\mathrm{Pe}}

@@

\usepackage{threeparttable}
\usepackage{longtable}

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{Sediment Interception by Emergent Stems Across Varying Patch Densities and Flows}

@@

%TODO
% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Suspended sediment collected by vegetation in marshes and wetlands contributes to vertical accretion, which can buffer against rising sea levels and preserve these ecosystems. Effective capture efficiency (ECE), a parameter relating the rate of sediment capture by submerged vegetation surfaces to their effective diameters, is a key variable in determining the significance of direct interception in elevation-change models. The factors governing ECE are not yet thoroughly understood in transitionally turbulent flows. Using a recirculating flume containing an array with a variable count of artificial cylindrical collectors, we conducted a set of 12 experiments at 3 flow speeds and 3 collector densities (plus control), plus 4 additional experiments where biofilm was allowed to accumulate. We determined that ECE decreases with increasing collector Reynolds number (study range: 66 to 200; p < 0.05 for 2 of 3 treatments) and collector density (solid volume fraction: 0.22\% to 1.17\%; p < 0.05 for 2 of 3 treatments). Adding biofilm increased ECE in all cases (improvement to 1.53--7.15 times corresponding experiment without biofilm). By combining our data with those of the most similar study, we also present a preliminary model quantitatively assessing the effect of collector density on ECE.}
